---
title: "Biomass_AmpEff_variation_func"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r dependencies}
library(here)
library(tidyverse)
library(binr)
library(lattice)
library(viridisLite)
```

The goal of following function (TechDrops.vs.CountsAmps) is to see how the frequency of dropouts among technical reps varies as a function of amplification efficiency and original biomass or count. 

Single argument is an Input dataframe with the following columns: 
- tech_1, tech_2, and tech_3 with read counts from each technical replicate
- abundance, with manual counts or other information on original template amount 
- amp_ef, with amplication efficiency for each asv-sample (each row, essentially)

```{r tech variability as function of amp efficiency and biomass}
TechDrops.vs.CountsAmps <- function(input_df, countBreaks, ampBreaks){
  
  total_binned_df <- data.frame(input_df, abundance_bin=cut(input_df$abundance, na.rm =T, breaks = countBreaks),  ampef_bin = cut(input_df$amp_ef, na.rm= T, breaks = ampBreaks))  # sort abundance and amp efficiency into quantiles with all input data
  
  unique_bins <- unique(total_binned_df[,c("abundance_bin","ampef_bin" )]) # all unique amp bin and abundance bin combinations
  unique_bins <- unique_bins[complete.cases(unique_bins), ]
  
  list <- replicate(length(unique_bins[,1]), rep(NA, length=nrow(total_binned_df)), simplify = FALSE) # list of empty vectors (length of input dataframe), one vector for all the row indices corresponding to an unique amp&abundance bin combination
  
  propdrops <- unique_bins %>% mutate(dropout_proportion = rep(NA, length(unique_bins[,1]))) # where proportion of dropouts will go for each unique amp&abundance bin combination
  
  for(i in 1:length(unique_bins[,1])){
    bin <- as.numeric(unique_bins[i,]) # get vector of numeric factor levels of that abundance/amp bin (so c([1,3], [0.1,0.3]) would be c(1,1))
    
    # Gets index values in input dataframe corresponding to a the numeric factor levels for given abundance/amp bin 
    bin_levels_list <- lapply(total_binned_df[,c('abundance_bin','ampef_bin')], as.numeric)
    bin_indices <- which(bin_levels_list$abundance_bin == bin[1] & bin_levels_list$ampef_bin == bin[2])
    
    specific_bin_df <- total_binned_df[bin_indices,] # subset data to only include given abundance/amp bin, get prop of dropouts
    
    dropouts <- c(which(specific_bin_df$tech_1 == 0),
                  which(specific_bin_df$tech_2 == 0),
                  which(specific_bin_df$tech_3 == 0)) # need to filter out tech reps with all zeros
    propdrops[i,3] <- length(unique(dropouts)) / nrow(specific_bin_df)
    print(length(unique(dropouts)) / nrow(specific_bin_df))
  }
  
  
  print(unique_bins)
  # just adding numerical factor levels to the bin factor levels
  outputdf <- cbind(propdrops, unique(cbind(as.numeric(propdrops$abundance_bin), as.numeric(propdrops$ampef_bin))))
  
  # plot 
  
  print(levelplot(dropout_proportion ~ `2`*`1`, # "2" is amp, "1" is abundance
                  scales=list(x= list(at=sort(unique(outputdf$`2`)), labels = c("Low")),
                              y= list(at=sort(unique(outputdf$`1`)), labels = c("Low"))),
                  colorkey= list(title= "Proportion of Dropouts"),
                  data=outputdf,
                  xlab="amplification efficiency bins",
                  ylab = "abundance bins",
                  main="",
                  col.regions = viridis(100))) 
  
  return(outputdf)
}

```

```{r run function w dataset}
CalCofi_mock_complete <- read.csv(here("special_issue_manuscript","Data", "complete_cal_props_long.csv"))

# make ID column for each "asv"
CalCofi_mock_complete_IDrow <- CalCofi_mock_complete %>% unite(fullID, c(bio_rep,ID_mifish, ID_microscopy, ext_rep), sep=".") 
CalCofi_mock_complete_IDrow$fullID <- str_sub(CalCofi_mock_complete_IDrow$fullID, 2, -3)

# pivot wide
CalCofi_wide_reads_counts <- CalCofi_mock_complete_IDrow %>%
  select(fullID,tech_rep, larval_counts, mean_log_alpha, mifish_reads) %>% 
  pivot_wider(names_from = tech_rep, values_from = mifish_reads) %>% 
  select(-fullID)

CalCofi_mock_complete <- CalCofi_wide_reads_counts %>% rename(c("amp_ef" = mean_log_alpha ,"abundance" = larval_counts, "tech_1" = "1", "tech_2" = "2", "tech_3" = "3"))

test<- CalCofi_mock_complete[- c(which(CalCofi_mock_complete$"tech_1" == 0 & CalCofi_mock_complete$"tech_2" == 0, CalCofi_mock_complete$"tech_3" == 0)),] 


output <- TechDrops.vs.CountsAmps(input_df = test, countBreaks = 2, ampBreaks = 2)
```

