---
title: "Spread among technical replicates based on # of detections, # of reads, or taxon"
author: "Ramon Gallego, Helen Casendino, Kai Vennemann, Joe Duprey"
date: "12/29/2021"
output: html_document
---

```{r setup & dependencies, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(ggplot2)
```

# Code Description

This script takes as input raw read data from Gallego et al. (2020) to explain variation among technical replicates (as a function of detection #, read #, and taxa). 
1. Are agreeing technical replicates (3 detections) *less precise* with more rare things?
Subset asvs (site-date-bottle-hash) with detection in each PCR replicate (no dropouts), and calculate the CV for each set of technical replicates. Plot how CV tracks with log(mean(nReads)) and taxa. 
2. How does the proportion of dropouts (<3 detections) change with taxa?
Subset asvs (site-date-bottle-hash) with <3 detections. For a given taxon, calculate the percentage of that taxon's occurences with 1 or 2 PCR dropouts. Additionally, how does the ratio of 1 dropout/2 dropouts change with remaining nReads value? 

# Read in Data 

```{r read in data}
raw_reads <- read_csv(here("special_issue_manuscript/data/gallego_all_data.csv"))
hash_annotated <- read_csv(here("special_issue_manuscript/data/gallego_hash_annotated.csv"))
```

# Subset of (3) detections

```{r Creating df with annotations and CV}

wide_reads <- raw_reads %>%
  mutate(bottle = paste(site, bio, sep = "")) %>%
  dplyr::select(tech, hash, reads, bottle) %>%
  pivot_wider(names_from = tech, values_from = reads, values_fill = 0) %>% # fill non-detections with zero, then data to longform
  pivot_longer(cols = c("1", "2", "3"), names_to = "tech", values_to = "reads") %>%
  group_by(hash, bottle) %>%
  mutate(CV = sd(reads) / mean(reads)) %>% # calculate CV
  mutate(mean_reads = mean(reads)) %>%
  mutate(log_mean_reads = log(mean_reads)) %>%
  pivot_wider(names_from = tech, values_from = reads) %>%
  rename("Hash" = "hash")

# join read counts and hashes with hashes and taxa labels
wide_reads_annotated <- left_join(wide_reads, hash_annotated, by = "Hash")

```

```{r Subset 3 detections}
dropouts <- c(which(wide_reads_annotated$`1` == 0),
              which(wide_reads_annotated$`2` == 0),
              which(wide_reads_annotated$`3` == 0))
reads_annotated_no_drops <- wide_reads_annotated[-unique(dropouts),]
```

```{r Plots}

# log_mean_reads vs. CV
ggplot(reads_annotated_no_drops, aes(x = log_mean_reads, y = CV)) +
  geom_point() + 
  scale_x_discrete(guide = guide_axis(angle = 90))

# taxa vs. CV
# For now, let's go with these taxa to get a range of prevalence (needs work)
king <- c("Metazoa", "Rhodophyta", "Fungi","Dinoflagellates")

kingdom_subset_indices <- which(reads_annotated_no_drops$kingdom %in% king)

reads_annotated_kingdom_subset <- reads_annotated_no_drops[kingdom_subset_indices,]

ggplot(reads_annotated_kingdom_subset, aes(x = kingdom, y = CV)) +
  geom_boxplot() +
  scale_x_discrete(limits = king, guide = guide_axis(angle = 90))
```

# Exploring (<3) detections

```{r Per taxon, percentage of dropouts; plot}
prop.drops <- function(wide_df, kingdoms){

proportion_drops <- data.frame(kingdom = kingdoms, proportionDrops = NA)
for(i in 1:length(kingdoms)){
  df <- wide_df[wide_df$kingdom == kingdoms[i],]

  dropouts <- c(which(df$`1` == 0),
              which(df$`2` == 0),
              which(df$`3` == 0))
 proportion_drops[i,2] <- length(unique(dropouts)) / nrow(df)
}
  return(proportion_drops)
}

output_df <- prop.drops(wide_reads_annotated, c("Metazoa", "Rhodophyta", "Fungi","Dinoflagellates"))

ggplot(output_df, aes(x = kingdom, y = proportionDrops)) +
  geom_bar(stat="identity") + 
   scale_x_discrete(limits = kingdoms, guide = guide_axis(angle = 90)) 
```

We also know that one-detection asvs are linked to lower mean reads than two- and three-detection asvs. 

