---
title: "10212020_satellite_data_rerddapXtracto"
author: "Zack Gold"
date: "9/28/2021"
output: html_document
---

#Load Packages
```{r}
library(ncdf4) 
library(parsedate)
library(plotdap)
library(rerddap)
library(sp)
library(rerddapXtracto)
library(devtools)
library(rerddapXtracto)
library(tidyverse)
library(gganimate)
library(ggplot2)
library(plotdap)
library(here)
```

# Example 1 - CALCOFI

### SST
```{r}
require("rerddap")
## base URL does not need to given because it is the default one
dataInfo <- info('erdMBchla1day')

urlBase <- "https://coastwatch.pfeg.noaa.gov/erddap/"
parameter <- "sea_surface_temperature"

dataInfo <- info('erdPH53sstdmday', url = urlBase)
```
### Load in Site Sampling Data
Brief background, there are 4 stations sampled in the Spring 1996, 1998-2019. 
Hoping to obtain monthly data across the entire sampling time span, specifically to see if SST/Chla is a predictor variable for ichthyoplankton communities derived from eDNA data. Particularly interested in comparing 1 month, 2 month, and annual SST values.
```{r}
calcofi_sampling <-
  read.csv(here("data", "CALCOFI_date_loc_data.csv"))
  
  #Get monthly time data
  calcofi_sampling %>%
  filter(.,
  date %in% c("1996-04-16", "1996-04-28", "1996-04-23", "1996-04-17")) %>%
  dplyr::select(-date) -> date_template
  
  seq(as.Date("1995-04-01"), by = "month", length.out = 290) -> all_dates
  all_dates %>%  as_tibble() %>%
  dplyr::select(date = value) %>%
  merge(date_template) -> full_calcofi_sampling
  
  full_calcofi_sampling$date <-
  as.Date(full_calcofi_sampling$date, "%Y-%m-%d")
  full_calcofi_sampling %>%
  mutate(., lon = lon + 360) -> full_calcofi_sampling
  
  #Chla Satellite data only available 1997 and onwards
  
  #Get monthly time data
  seq(as.Date("1997-09-04"), by = "month", length.out = 270) -> all_dates_2
  all_dates_2 %>%  as_tibble() %>%
  dplyr::select(date = value) %>%
  merge(date_template) -> small_calcofi_sampling
  
  
  small_calcofi_sampling$date <-
  as.Date(small_calcofi_sampling$date, "%Y-%m-%d")
  small_calcofi_sampling %>%
  mutate(., lon = lon + 360) -> small_calcofi_sampling
  
```


```{r}
require("rerddap")
require("rerddapXtracto")

# First we will copy the calcofi_sampling data into a variable
xpos <- full_calcofi_sampling$lon
ypos <- full_calcofi_sampling$lat
tpos <- full_calcofi_sampling$date
zpos <- rep(0., length(xpos))
swsstInfo <- rerddap::info('erdPH53sstdmday', url = urlBase)

#This code works, but takes a while to run
# swsst1 <- rxtracto(swsstInfo, parameter = parameter, xcoord = xpos, ycoord = ypos, tcoord = tpos, xlen = .2, ylen = .2, progress_bar = TRUE)
# saveRDS(swsst1, file=here("data","sst_data.RDS"))

#Load
swsst1 <- readRDS(file = here("data", "sst_data.RDS"))
```

```{r}
require("ggplot2")
require("plotdap")

#Annimation Does not do what I hoped/expected
myPlot_1 <-
plotTrack(
swsst1,
xpos,
ypos,
tpos,
plotColor = 'thermal',
animate = TRUE,
cumulative = TRUE
)

```


#Example 3 CALCOFI
Plot entire Southern California Region.
### SST


```{r}
require("rerddap")
require("rerddapXtracto")
parameter <- "sea_surface_temperature"

xpos <- c(-122 + 360, -116 + 360)
ypos <- c(29, 35)
time_save <-
seq(as.Date("1995-04-01"), by = "year", length.out = 25)
tpos <- c(min(time_save), max(time_save))
sstInfo <- rerddap::info('erdPH53sstdmday', url = urlBase)

#Load data to save time
VIIRS <-
rxtracto_3D(
sstInfo,
parameter = parameter,
xcoord = xpos,
ycoord = ypos,
tcoord = tpos
)

saveRDS(VIIRS, file = here("data/sst_region_data.RDS"))

#Load
VIIRS <- readRDS(file = here("data/sst_region_data.RDS"))

```

```{r}
require("gganimate")
require("ggplot2")
require("plotdap")

seq(1, 290, 12) -> slicer #Only take the month of CalCOFI sampling in Spring instead of all 288 months
sanctchl1 <- VIIRS
sanctchl1$sea_surface_temperature <-
sanctchl1$sea_surface_temperature[, , slicer]
sanctchl1$time <- sanctchl1$time[slicer]

sst_Plot2 <-
plotBBox(
sanctchl1,
plotColor = 'thermal',
time = identity,
animate = TRUE,
maxpixels = 30000
)

sst_Plot2
gganimate::anim_save(here("data", "sst_Plot.gif"))

```