---
title: "Quality Control for OA Data"
author: "Helen Casendino, Joe Duprey, Ramon Gallego, Kai Vennemann"
date: "10/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quality Control for OA Data

This code takes our raw data and saves it in a standard form that is reusable across many of our functions. It creates two files: all_data.csv and all_data_clean.csv. The latter file is similar to the former but has outliers removed from it.

```{r dependencies}
library(tidyverse)
```

## Step 1: Load raw ASV data.

We use our OA data.

```{r load data}
raw_ASVs <- read_csv("../input/raw_ASVs.csv")
```

## Step 2: Save data as CSV in standard form.

Our standard form is a tidy data set with the following columns: seq_run, site, bio, tech, hash, reads. In order, these represent the DNA sequencing run number, location and date of sampling, biological replicate (bottle), technical replicate (PCR), species (encoded here as a hash), and number of occurrences of that species.

```{r save data in standard form}
all_data <- raw_ASVs %>%
  rename(seq_run = Miseq_run, hash = Hash, reads = nReads) %>% # Consistent naming
  filter(str_detect(sample, "Ostrich", TRUE)) %>% # Remove ostrich samples
  filter(str_detect(sample, "Kangaroo", TRUE)) %>% # Remove kangaroo samples
  filter(str_detect(sample, "K+", TRUE)) %>% # Remove negative control samples
  filter(str_detect(sample, "k+", TRUE)) %>%
  mutate(sample = gsub("_", "", sample)) %>%
  separate(col = sample, into = c("sample", "tech"), sep = "[.]", remove = FALSE) %>%
  mutate(tech = as.integer(tech)) %>%
  mutate(site = gsub(".{1}$", "", sample)) %>% # Separate site and bottle info from each other
  mutate(bio = str_sub(sample, -1)) %>%
  dplyr::select(seq_run, site, bio, tech, hash, reads)

# Check for repeats
repeats <- all_data %>%
  group_by(site, bio, tech, hash, reads) %>%
  filter(n() > 1)
if (nrow(repeats) > 0) warning("all_data has repeats!")

write_csv(all_data, "../data/all_data.csv")
```

## Step 3: Clean data by removing really low reads.

The following code is based on Denoising.all.runs.Rmd in the OA paper. We fit the reads from all of the technical replicates to a normal distribution. Then we remove low reads outside of a 92.5% CI. 

```{r fit reads to normal dist}
reads_by_sample <- all_data %>%
  mutate(sample = paste(site, bio, tech, sep = "")) %>%
  dplyr::select(sample, reads) %>%
  group_by(sample) %>%
  summarise(tot = sum(reads)) # Sum of reads for each PCR replicate

# Visualize total read distribution
hist(reads_by_sample$tot, breaks = 80)

normparams <- MASS::fitdistr(reads_by_sample$tot, "normal")$estimate

# Reads outside of our specified CI
outliers <- reads_by_sample %>%
  mutate(reads_by_sample, prob = pnorm(tot, normparams[1], normparams[2])) %>%
  filter(prob < 0.075 & tot < normparams[1])
```

Here we remove the outliers we determined above and save the clean data as a CSV.

```{r create all_data_clean.csv}
all_data_clean <- all_data %>%
  mutate(sample = paste(site, bio, tech, sep = "")) %>%
  filter(!sample %in% outliers$sample) %>%
  dplyr::select(-sample)

write_csv(all_data_clean, "../data/all_data_clean.csv")
```
