---
title: "rareSp.dissimilarity"
author: "Helen Casendino"
date: "7/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r Dependencies, include=FALSE}
library(tidyverse)
library(vegan)
```

This code will look at the impact of rare species on dissimilarity between PCR replicates. We'll start with PCR.proportions.clean.csv, which has the proportions of each hash by PCR replicate (ONLY triplicate PCR samples) and really low read PCR replicates are removed. 

```{r load in data}
full.proportion.reads <- read.csv('../data/PCR.proportions.clean.csv')
```

Below is my start to making the subsetting loop with BCD centroids for each sample. So, for each sample, we'll have a rank of most common hashes to least common. We will subset this with varying exclusion proportions (ie, removing bottom 1% of hashes, bottom 5%). For each subset, calculate the centroid of 3 pairwise bray curtis dissimilarities: PCRs 1&2, 2%3 and 1&3.   

```{r Function that calculates centroids by sample across rare sp subsets}

# df is a data frame with 6 columns: Miseq_run, bio, Hash, PCR1_prop, PCR2_prop, PCR3_prop
# exclusionProp is a vector of all the subsetting proportions you want. A vector of (0.01, 0.05) means you'll get the PCR centroid for when 1% of the lowest proportion samples are excluded, and when 5% of lowest prop samples are excluded.
# bioVector is a vector of all biological sample IDs in your dataset.

PCRcentroid_bySample <- function(df, exclusionProp, bioVector){
   
  mat <- matrix(NA, nrow = length(exclusionProp), ncol = length(bioVector))
  
  for(j in 1:length(bioVector)){
    
      bioSample <- bioVector[j]
      bioProps_Hashes <- df[df$bio == bioSample,] # to get all PCR props from that one sample
      
      meanProp.acrossPCR <- bioProps %>% group_by(Hash) %>% summarise(meanProp.PCR = mean(c(PCR1_prop, PCR2_prop,PCR3_prop))) %>% arrange(desc(meanProp.PCR)) # to get rarity order
      
      for(i in 1:length(exclusionProp)){
       cutoff <- round(length(meanProp.acrossPCR$Hash) - length(meanProp.acrossPCR$Hash)*exclusionProp[i])
        sub.tib <- meanProp.acrossPCR[1:cutoff,]
        sub.tib <- bioProps_Hashes %>% filter(Hash %in% sub.tib$Hash) %>% select(!c(Hash, bio,Miseq_run))
       
         flip.tib <- t(sub.tib)
         dis <- vegdist(flip.tib)
        grouping <- factor(rep(1,3), labels=bioSample)
       
         centroid<- betadisper(dis, group = grouping)
        centroid$distances # I definitely did this wr
        
        #### GET CENTROID HERE SOMEHOW
        mat[i,j] <- mean(centroid$distances)
      }
    
    print(j)  # progress bar
  }
  return(mat)
}

# returns a matrix with biological replicate as columns and exclusion proportion as rows. 

```

Now let's visualize it with our data

```{r Organize function matrix output}

exclvec<- seq(0,0.2, 0.01)
centroids <- PCRcentroid_bySample(full.proportion.reads, exclvec, unique(full.proportion.reads$bio))

mean.centroidDists <- rowMeans(as.data.frame(centroids)) # average distances to centroid of PCR reps for all exclusion proportion values across biological replicates

cutoff.centroid.df <- data.frame(exclusion.prop = exclvec, cent.distance = mean.centroidDists) # combines with exclusion prop vector
```


```{r Plot data}
plot(cutoff.centroid.df$exclusion.prop, cutoff.centroid.df$cent.distance, pch = 19, xlab = "% Rare Species Excluded", ylab = "Distance to Centroid")

```


