---
title: "Community Level PCR & Bottle Variation - Bray Curtis"
author: "Helen Casendino, Joe Duprey"
date: "9/28/2021"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r Dependencies, include=FALSE}
library(tidyverse)
library(vegan)
library(stringi)
```

This code functions similarly to centroid.comm.var.Rmd in that it creates distributions of variability among technical (PCR) and biological (bottle) replicates at the community level, using the OA data with triplicate PCRs. It functions differently in a few ways: 
     
      1) Instead of calculating the distance to the centroid for each bottle or sampling event, it harvests all pairwise bray-curtis distances (BCDs) from the vegdist() function's output.
      2) For the biological replicate level (distance between bottles within sampling event), there is NO AVERAGING ACROSS PCRs (i.e., we are collecting pairwise BC distances among A.1,A.2,A.3,B.1,B.2, etc.).
      3) To avoid tabulating intra-bottle PCR variation when looking at the biological replicate level, we will only harvest pairwise distances between communities in *different* bottles...ie, take the pairwise distance between A.1 and B.1,B.2,B.3, but not between A.1 and A.2,A.3
      
## Step 1: Read in the data
Note that PCR.proportions.clean.csv is a data frame with the following structure:
        bio                                     Hash    PCR1_prop    PCR2_prop  PCR3_prop
1 CP201706A 0027eb8cc37fef2af5a828e5c7f186d1a639ec73 0.0047199782 0.0000000000 0.00000000
2 CP201706A 00616a421c92928530cd4f6653722e00616d21e7 0.0473813198 0.0134351405 0.02373608

```{r load in data}
trip.proportion.reads <- read.csv('../../data/PCR.proportions.clean.csv')
```
      
## Step 2: PCR Variation
Create and run a function that calculates and plots PCR pairwise BCDs by biological replicate. Our example uses 3 pairwise comparisons (by choosing pairs from among PCR1_prop, PCR2_prop, PCR3_prop).

```{r create & run function to calculate BCDs by bottle; Plot}
# Function arguments: 
# a) df is a data frame with columns for each PCR run and a column for bottle ID
# b) bottles is a vector of all biological sample IDs in your data set
# c) PCR.cols is a vector of column indices for all the PCR reactions
PCR.bcd <- function(df, bottles, PCR.cols) {
  bcds <- c("start") # will hold all bray curt distances (BCDs)
  
  for(j in 1:length(bottles)) {
    bottle.hashes <- t(df[df$bio == bottles[j], PCR.cols]) # to get all PCR props from that one sample
    # vegdist() requires us to take the transpose
    dis <- vegdist(bottle.hashes)
    bcds <- c(bcds, as.vector(dis)) # takes all values of the vegdist output
    # print(as.vector(dis))
  }
  # plot
   his <- hist((as.numeric(bcds[-1])), col = viridis::plasma(3,0.4,0.7), main = "PCR Variation", xlab = "Bray-Curtis Pairwise Distances")
  
    return(as.numeric(bcds[-1])) 
    return(his)
}

PCR.output <- PCR.bcd(trip.proportion.reads,  unique(trip.proportion.reads$bio), 3:5)
# mean = 0.334
```

## Step 3: Bottle Variation
As a reminder, we are NOT averaging across PCRs.

```{r data wrangling, not taking PCR mean}
long.props <- trip.proportion.reads %>% 
  mutate(sample = gsub('.{1}$', '', bio)) %>% 
  mutate(bottle = stri_sub(bio,-1)) %>% 
  pivot_longer(cols = c(PCR1_prop,PCR2_prop,PCR3_prop), names_to = "PCR", values_to = "Proportion")

get.PCRn <- long.props %>% 
  mutate("PCRn" = stri_sub(PCR,4,-6)) %>% 
  select(-c(bio,PCR))

nested.props <- get.PCRn %>% 
  unite(bottle, PCRn, col = "bio.PCR", sep = ".") %>% 
  nest(data = c(bio.PCR,Hash,Proportion))

for(i in 1:length(nested.props$sample)) {
  
  nested.props$data[[i]] <- nested.props$data[[i]] %>% 
    group_by(bio.PCR,Hash) %>% 
    summarise(Proportion = mean(Proportion)) %>%  
    pivot_wider(names_from = bio.PCR, values_from= Proportion, values_fill = 0) 
  
  print(i)
}
```


```{r create & run function to calculate BCDs by sampling event; Plot}
# list = list of data frames corresponding to each sampling event. Within each df, columns correspond to bioPCR combos (A1, B3, C2 etc) and rows correspond to each Hashes
# SampleVector is vector of sampling events (PO20170311, etc)
BOTTLE.bcd <- function(list, SampleVector){
  bcds <- "start"
 
  # Getting rid of samples with only one bottle
       onebottle <- rep(NA, length=100) # we will put the "list" indices of replicates with only one bottle here
  
       for(i in 1:length(SampleVector)){
         a <- list[[i]]
          if(((ncol(a) -1)/3) == 1){ # replicates with only 1 bottle 
           onebottle[i] <- i # save index
             SampleVector <- SampleVector[-i] # remove index from sample vector
  } 
  }
        onebottle <- onebottle[!is.na(onebottle)] # remove NAs
        list <- list[-onebottle] # remove the one bottle elements from list
   
  for(j in 1:length(SampleVector)){
    
    print(j)
     
    df <- list[[j]]
    sub.tib <- df %>% select(!c(Hash))
    flip.tib <- t(sub.tib)
    dis <- vegdist(flip.tib) 
    dis.mat <- as.matrix(dis) 
    
    # PRINT DISTANCE MATRICES BEFORE REMOVAL <- can remove later // joe getting up to speed
    print(SampleVector[j])
    print(dis.mat)
    
    # getting rid of intra-bottle pairwise comparisons...
    # remove all cells corresponding to distances within the same bottle
    tryCatch({
     dis.mat[1:ncol(dis.mat),1:3] <- NA
    dis.mat[4:ncol(dis.mat),4:6]  <- NA
    dis.mat[7:ncol(dis.mat),7:9] <- NA
    dis.mat[10:ncol(dis.mat),10:12] <- NA
    dis.mat[13:ncol(dis.mat),13:15] <- NA
    dis.mat[16:ncol(dis.mat),16:18] <- NA
     if (j==j) stop("subscript out of bounds")
    }, error=function(e){})
    
    # PRINT DISTANCE MATRICES AFTER REMOVAL <- can remove later // joe getting up to speed
    print(dis.mat)

    bcd.vector <- as.numeric(na.omit(as.vector(dis.mat))) # remove intra-bottle comparisons
     bcds <- c(bcds, bcd.vector) # all bray curtis dists for that sample
  }
        
  #plot
       his <- hist((as.numeric(bcds[-1])), col = viridis::viridis(3,0.4,0.7), main = "Bottle Variation", xlab = "Bray-Curtis Pairwise Distances")
  
 return(as.numeric(bcds[-1])) 
 return(his)
}

bottle.output <- BOTTLE.bcd(nested.props$data, nested.props$sample)
# mean = 0.431
```

      