---
title: "Community Level PCR & Bottle Variation"
author: "Helen Casendino"
date: "8/30/2021"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r Dependencies, include=FALSE}
library(tidyverse)
library(vegan)
library(gridExtra)
library(stringi)
```

## Step 1: Read in the data. 

 - trip.proportion.reads = the proportions of each hash by PCR replicate (ONLY triplicate PCR samples) and biological samples with really 1 low read PCR replicate are removed

```{r load in data}
trip.proportion.reads <- read.csv('../../data/PCR.proportions.clean.csv')

dup.proportion.reads <- read.csv('../../data/PCR.duplic.proportions.csv')
```

## Step 2: PCR Variation: Create and run function to get mean distance to centroid for 3 different PCR pairwise comparisons, by biological replicate

```{r create & run function to calculate centroids by bottle; Plot}
# Function calls: 
# a) df is a data frame with columns for each PCR run and a column for bottle ID
# b) bottles is a vector of all biological sample IDs in your dataset.
# c) PCR.cols is vector of column indices for each PCR rxn 

PCRcentroids <- function(df, bottles, PCR.cols){
  
  dist.Centroid <-  rep(NA, length(bottles))
  
  for(j in 1:length(bottles)){

    bottle.hashes <- t(df[df$bio == bottles[j],PCR.cols]) # to get all PCR props from that one sample
    dis <- vegdist(bottle.hashes)
    grouping <- factor(rep(1,3), labels=bottles[j])
    centroid<- betadisper(dis, group = grouping)
    dist.Centroid[j] <- mean(centroid$distances)
    print(j) # progress bar
  }
  return(dist.Centroid)
}

PCR.output <- PCRcentroids(trip.proportion.reads,  unique(trip.proportion.reads$bio), 3:5)

hist(PCR.output, col = viridis::plasma(3,0.4,0.7), main = "PCR Variation", xlab = "Distance to Centroid (BC Pairwise Distances)")
# text(0.4, 30,"mean = 0.198")
```

## Bio-PCR Centroid Distances: Bottle Variation

- Extra code calculating same thing as above (distance to centroids), just loops over sampling events (not bottles). Here, we will take the average across PCR replicates for each Hash. 
- we need to get averages of Hashes across PCRs, nest data within samples, and widen it

```{r data wrangling}
#1) Remove PCR columns (make long), Split into sample and bottle
long.props <- trip.proportion.reads %>% 
  mutate(sample = gsub('.{1}$', '', bio)) %>% 
  mutate(bottle = stri_sub(bio,-1)) %>% 
  pivot_longer(cols = c(PCR1_prop,PCR2_prop,PCR3_prop), names_to = "PCR", values_to = "Proportion")
#2) Get average proportion across PCRs for each bottle/hash combo
get.PCRn <- long.props %>% 
  mutate("PCRn" = stri_sub(PCR,4,-6)) %>% 
  select(-c(bio, PCR)) %>%
  unite(sample, bottle, col = biorep, sep = ".", remove=FALSE) %>% 
  group_by(biorep, Hash) %>%
  summarise(meanprop=mean(Proportion))
#3) separate into sample and bottle columns
get.PCRn <-get.PCRn %>%
  separate(col=biorep, into=c('sample','bio'), sep='[.]', remove=TRUE)
#4) nest data
nested.props <- get.PCRn %>% 
  nest(data = c(bio, Hash, meanprop))
#5) Within each sample, widen so that each column corresponds to a bottle
for(i in 1:length(nested.props$sample)) {
  nested.props$data[[i]] <- nested.props$data[[i]] %>% 
    pivot_wider(names_from = bio, values_from=meanprop, values_fill = 0) 
    print(i)
}
```

```{r create & run function to calculate centroids by sample; Plot}
# list = list of data frames corresponding to each sampling event. Within each df, columns correspond to bioPCR combos (A1, B3, C2 etc) and rows correspond to each Hashes
# SampleVector is vector of sampling events (PO20170311, etc)

BOTTLEcentroids <- function(list, SampleVector){
  
  centroid.dists <- rep(NA, length(SampleVector))
  
  for(j in 1:length(SampleVector)){
    df <- list[[j]]
    sub.tib <- df %>% select(!c(Hash))
    nBIOPCR <- ncol(sub.tib)
    flip.tib <- t(sub.tib)
    
    dis <- vegdist(flip.tib)
    grouping <- factor(rep(1,nBIOPCR), labels=SampleVector[j])
    centroid<- betadisper(dis, group = grouping)
    centroid.dists[j] <- mean(centroid$distances)
    print(j)
  }
  return(centroid.dists)
}

bottle.output <- BOTTLEcentroids(nested.props$data, nested.props$sample)

hist(bottle.output, col = viridis::viridis(3,0.4,0.7), main = "Bottle Variation", xlab = "Distance to Centroid (BC Pairwise Distances)")
# text(0.4, 12,"mean = 0.229")
```

#########
I ALSO WANT TO SEE HOW ANSWER TO BOTTLE VARIATION CHANGES WITHOUT AVERAGING ACROSS PCRS
###############

```{r data wrangling, not taking PCR mean}
long.props <- trip.proportion.reads %>% 
  mutate(sample = gsub('.{1}$', '', bio)) %>% 
  mutate(bottle = stri_sub(bio,-1)) %>% 
  pivot_longer(cols = c(PCR1_prop,PCR2_prop,PCR3_prop), names_to = "PCR", values_to = "Proportion")

get.PCRn <- long.props %>% 
  mutate("PCRn" = stri_sub(PCR,4,-6)) %>% 
  select(-c(bio,PCR))

nested.props <- get.PCRn %>% 
  unite(bottle, PCRn, col = "bio.PCR", sep = ".") %>% 
  nest(data = c(bio.PCR,Hash,Proportion))

for(i in 1:length(nested.props$sample)) {
  
  nested.props$data[[i]] <- nested.props$data[[i]] %>% 
    group_by(bio.PCR,Hash) %>% 
    summarise(Proportion = mean(Proportion)) %>%  
    pivot_wider(names_from = bio.PCR, values_from= Proportion, values_fill = 0) 
  
  print(i)
}
```

```{r run BOTTLEcentroids function again}
bottle.output.bioPCR <- BOTTLEcentroids(nested.props$data, nested.props$sample)
hist(bottle.output.bioPCR, col = viridis::viridis(3,0.4,0.7), main = "Bottle Variation (no PCR averaging)", xlab = "Distance to Centroid (BC Pairwise Distances)")
mean(bottle.output.bioPCR) # mean is still 0.299...so higher than the earlier bottle var estimate
#text(0.5, 12,"mean = 0.299")
```

