---
title: "Exploring various methods of assessing dissimilarity between communities within different PCR replicates"
author: "Joe Duprey, Helen Casendino"
date: "9/5/2021"
output: html_document
---

This document looks at using Bray Curtis Dissimilarity, Jaccard methods, and Beta Diversity to assess dissimilarity between communities within different PCR replicate combinations. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r Dependencies, include=FALSE}
library(tidyverse)
library(vegan)
library(reshape)
library(stringi)
```

Our broad goal here is to try 3 different methods — Jaccard, Bray-Curtis, and Beta Diversity — to quantify dissimilarity in the "communities" of PCR replicates 1, 2 and 3. 

We start with clean.tech.table.csv (from divide.reps_calc.eDNA.index.Rmd), which includes triplicates, not duplicates. Joe and I think the eDNA index function used isn't perfect yet, but we'll work with it for now. 

```{r Load data}
trip.proportion.reads <- read.csv('../../data/PCR.proportions.clean.csv')
```

## Step 1: Try Bray Curtis (outdated?) 

First, convert to community matrix.

```{r data to matrix}
prop.reads <- trip.proportion.reads[,4:6] # only eDNA proportion values
prop.mat<- t(prop.reads) # columns = Hashes, rows = PCR #, values = nReads proportions
```



```{r BCD between PCR runs across samples}
vegdist(prop.mat, method="bray", na.rm = T)
```

All yielding pretty similar BC values (~0.41). But I'm worried that we might be glossing over some variability between bottles (ie, variation b/w PCR rep 1, 2 and 3 might be different based on the sample). Let's try getting PCR BC dissimilarity by sample. 

```{r BCD between PCR runs between samples}

 #bioVector = list of unique biological replicate IDs, indexCols = col numbers of eDNA proportion values (by PCR #). df needs to have 3 columns corresponding to eDNA proportions of different PCR runs

bray_by_sample <- function(df, bioVector, indexCols){  
  
    bray.mat <- matrix(NA, nrow = length(bioVector), ncol = 3, dimnames = list(bioVector, c(1.2,2.3,1.3))) # different PCR rep combinations
  
  for(i in 1:length(bioVector)){
    sample.tib <- df[df$bio == bioVector[i],indexCols]
    bray.tib <- t(sample.tib)
    bray.results <- as.matrix(vegdist(bray.tib, method="bray", na.rm = T))
    bray.mat[i,] <- c(bray.results[1,2], bray.results[3,2], bray.results[1,3]) # PCR 1&2, 2&3, 1&3
  }
    return(bray.mat)
}

bray_by_sample(trip.proportion.reads, unique(trip.proportion.reads$bio), c(4:6))

```

Also, I don't want to overfocus on variability between samples, because that's not what we're interested in. So maybe the chunk above is totally obsolete. However, I wonder if doing a permanova between the three *distributions* of BCDs (separated by sample) is more informative than just looking at 3 BCD values (for each PCR combination, across samples). ????

First, let's visualize how BCD distributions look in different PCR runs. 

```{r Visualize BCD distributions}

bray.output <-   bray_by_sample(trip.proportion.reads, unique(trip.proportion.reads$bio), c(4:6))

par(mfrow = c(1, 3))
hist(bray.output[,1], main="PCRs 1 & 2", xlab="BCD", ylim=c(0,60), col = "orange") 
hist(bray.output[,2], main="PCRs 2 & 3", xlab="BCD", ylim=c(0,60), col = "orange") 
hist(bray.output[,3], main="PCRs 1 & 3", xlab="BCD", ylim=c(0,60), col = "orange")
```

Now, let's compare distributions. We have to change the matrix around a bit.

```{r PERMANOVA on 3 BCD distributions}

bray.perm <- melt(as.data.frame(bray.output))
colnames(bray.perm) <- c("PCR.Combo", "BCD")
adonis(BCD ~ PCR.Combo,data = bray.perm, permutations = 9999)
```

Hm p value of 0.397. So at very first glance, reads don't differ that much between PCR replicates. (p < 0.05 = sig dissimilarity) 

*Notes on permanova + adonis:* 
- "Permutational multivariate analysis of variance (PERMANOVA) is a non-parametric multivariate statistical test. It is used to compare groups of objects and test the null hypothesis that the centroids and dispersion of the groups as defined by measure space are equivalent for all groups." 
- So we want to see how the DISSIMILARITY between samples in PCR Rep 1&2 compares to the DISSIMILARITY between samples in PCR Rep 1&3, etc. 
- adonis tests for homogeneity of dispersion among groups


## Step 2: Try Beta Diversity (looking at the number of unique hashes/taxa in samples between reps)

So beta diversity looks at total # of unique species in each PCR replicate. 
```{r Beta Div function for PCR reps & plot}
# again, betadiver wants PCR reps (equate to site) as rows, and Hashes (equate to species) as columns
# betadiver(help=T) lists methods...which one to use? I'll go with "sim" for now, Koleff et al. (2003) seem to think it's a good one.

beta.PCR <- function(df, bottles, PCR.cols){  
    betas <-  rep(NA, length(bottles))
   
      # proportion data needs to be in presence absence, convert proportion data to binary
  for(m in PCR.cols){
    df[which(df[,m] != 0),m] <- 1
  }
    
  for(j in 1:length(bottles)){
     beta.tib <- t(df[df$bio == bottles[j],PCR.cols]) 
    beta.dis <- betadiver(beta.tib, method = "sim")
    grouping <- factor(rep(1,3), labels=bottles[j])
    centroid<- betadisper(beta.dis, group = grouping)
    betas[j] <- mean(centroid$distances)
 #   print(j)
  }
    return(betas)
}

beta.PCR.output<- beta.PCR(trip.proportion.reads, unique(trip.proportion.reads$bio),3:5)

# Warning: In betadisper(beta.dis, group = grouping) :
#  some squared distances are negative and changed to zero
# Occured when j = c(31, 107, 175, 198,199,202,203,206)...not sure why this is happening, but I can just delete these rows for now
warnings <- c(31, 107, 175, 198,199,202,203,206)
beta.PCR.output <- beta.PCR.output[-warnings]

#plot
hist(beta.PCR.output, col = viridis::plasma(3,0.4,0.7), main = "PCR Variation", xlab = "Distance to Centroid (Beta Diversity)")
# text(0.15, 40,"mean = 0.285")
```

2) Getting beta div between bottles...need to nest data

```{r data wrangling, not taking PCR mean}
long.props <- trip.proportion.reads %>% 
  mutate(sample = gsub('.{1}$', '', bio)) %>% 
  mutate(bottle = stri_sub(bio,-1)) %>% 
  pivot_longer(cols = c(PCR1_prop,PCR2_prop,PCR3_prop), names_to = "PCR", values_to = "Proportion")

get.PCRn <- long.props %>% 
  mutate("PCRn" = stri_sub(PCR,4,-6)) %>% 
  select(-c(bio,PCR))

nested.props <- get.PCRn %>% 
  unite(bottle, PCRn, col = "bio.PCR", sep = ".") %>% 
  nest(data = c(bio.PCR,Hash,Proportion))

for(i in 1:length(nested.props$sample)) {
  
  nested.props$data[[i]] <- nested.props$data[[i]] %>% 
    group_by(bio.PCR,Hash) %>% 
    summarise(Proportion = mean(Proportion)) %>%  
    pivot_wider(names_from = bio.PCR, values_from= Proportion, values_fill = 0) 
  
  print(i)
}
```

3) Getting beta div between bottle.PCR runs 

```{r Beta between bottle+PCR runs within samples and visualize}

beta.Bottle <- function(list, SampleVector){
  
  centroid.dists <- rep(NA, length(SampleVector))
  
  for(j in 1:length(SampleVector)){
    df <- list[[j]]
    sub.tib <- df %>% select(!c(Hash))
    
     for(m in 1:ncol(sub.tib)){
    sub.tib[which(sub.tib[,m] != 0),m] <- 1
     }
    
    flip.tib <- t(sub.tib)
    beta.dis <- betadiver(flip.tib, method = "sim")
    grouping <- factor(rep(1,ncol(sub.tib)), labels=SampleVector[j])
    centroid<- betadisper(beta.dis, group = grouping)
    centroid.dists[j] <- mean(centroid$distances)
    print(j)
  }
  return(centroid.dists)
}

beta.bottle.output <- beta.Bottle(nested.props$data, nested.props$sample)

# gave same warning when j=c(11,22,35,60,68,69), so let's cut those for now. 
warnings <- c(31, 107, 175, 198,199,202,203,206)
beta.bottle.output <- beta.bottle.output[-warnings]

#visualize
hist(beta.bottle.output, col = viridis::viridis(3,0.4,0.7), main = "Bottle Variation (no PCR averaging)", xlab = "Distance to Centroid (Beta Diversity)")
#text(0.2, 12,"mean = 0.357")
```

## Step 3: Try Jaccard Similarity (bias towards rare sp)

Uses code similar to comm.level.var.Rmd
*Notes on jaccard* 
- "measures the similarity between two sets of data. It can range from 0 to 1. The higher the number, the more similar the two sets of data."
- vegdist jaccard (same as earlier BCD) wants taxa/hashes as columns...need data to be binary 

1) Getting jaccard index between PCR runs 

```{r Jaccard between PCR runs within bottles and visualize}

jaccard.PCR <- function(df, bottles, PCR.cols){  
  jaccards <-  rep(NA, length(bottles))
  
  # proportion data needs to be in presence absence, convert proportion data to binary
  for(m in PCR.cols){
    df[which(df[,m] != 0),m] <- 1
  }
  
  for(j in 1:length(bottles)){
    bottle.hashes <- t(df[df$bio == bottles[j],PCR.cols]) # to get all PCR props from that one sample
    dis <- vegdist(bottle.hashes, method="jaccard")
    grouping <- factor(rep(1,3), labels=bottles[j])
    centroid<- betadisper(dis, group = grouping)
    jaccards[j] <- mean(centroid$distances)
    print(j) # progress bar
  }
    return(jaccards)
}

jac.PCR.output <- jaccard.PCR(trip.proportion.reads, unique(trip.proportion.reads$bio), 3:5)

hist(jac.PCR.output, col = viridis::plasma(3,0.4,0.7), main = "PCR Variation", xlab = "Distance to Centroid (Jaccard Indices)")
# text(0.27, 50,"mean = 0.418")
```

2) Getting jaccard index between bottles...need to nest data

```{r data wrangling, not taking PCR mean}
long.props <- trip.proportion.reads %>% 
  mutate(sample = gsub('.{1}$', '', bio)) %>% 
  mutate(bottle = stri_sub(bio,-1)) %>% 
  pivot_longer(cols = c(PCR1_prop,PCR2_prop,PCR3_prop), names_to = "PCR", values_to = "Proportion")

get.PCRn <- long.props %>% 
  mutate("PCRn" = stri_sub(PCR,4,-6)) %>% 
  select(-c(bio,PCR))

nested.props <- get.PCRn %>% 
  unite(bottle, PCRn, col = "bio.PCR", sep = ".") %>% 
  nest(data = c(bio.PCR,Hash,Proportion))

for(i in 1:length(nested.props$sample)) {
  
  nested.props$data[[i]] <- nested.props$data[[i]] %>% 
    group_by(bio.PCR,Hash) %>% 
    summarise(Proportion = mean(Proportion)) %>%  
    pivot_wider(names_from = bio.PCR, values_from= Proportion, values_fill = 0) 
  
  print(i)
}
```

3) Getting jaccard index between bottle.PCR runs 

```{r Jaccard between bottle+PCR runs within samples and visualize}

jaccard.Bottle <- function(list, SampleVector){
  
  centroid.dists <- rep(NA, length(SampleVector))
  
  for(j in 1:length(SampleVector)){
    df <- list[[j]]
    sub.tib <- df %>% select(!c(Hash))
    
     # convert proportions to binary 
     for(m in 1:ncol(sub.tib)){
    sub.tib[which(sub.tib[,m] != 0),m] <- 1
     }
    
    flip.tib <- t(sub.tib)
    dis <- vegdist(flip.tib, method = "jaccard")
    grouping <- factor(rep(1,ncol(sub.tib)), labels=SampleVector[j])
    centroid<- betadisper(dis, group = grouping)
    centroid.dists[j] <- mean(centroid$distances)
    print(j)
  }
  return(centroid.dists)
}

jac.bottle.output <- jaccard.Bottle(nested.props$data, nested.props$sample)

#visualize
hist(jac.bottle.output, col = viridis::viridis(3,0.4,0.7), main = "Bottle Variation (no PCR averaging)", xlab = "Distance to Centroid (Jaccard Indices)")
#text(0.37, 20,"mean = 0.506")
```
