---
title: "Exploring various methods of assessing dissimilarity between communities within different PCR replicates"
author: "Joe Duprey, Helen Casendino"
date: "7/21/2021"
output: html_document
---

This document looks at using Bray Curtis Dissimilarity, Jaccard methods, and Beta Diversity to assess dissimilarity between communities within different PCR replicate combinations. I could also combine the 3 functions (bray_by_sample, beta_by_sample, jac_by_sample) into 1 function.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r Dependencies, include=FALSE}
library(tidyverse)
library(vegan)
library(reshape)
```

Our broad goal here is to try 3 different methods — Jaccard, Bray-Curtis, and Beta Diversity — to quantify dissimilarity in the "communities" of PCR replicates 1, 2 and 3. 

We start with clean.tech.table.csv (from divide.reps_calc.eDNA.index.Rmd), which includes triplicates, not duplicates. Joe and I think the eDNA index function used isn't perfect yet, but we'll work with it for now. 

```{r Load data}
full.proportion.reads <- read.csv("../data/PCR.proportions.csv")
```

## Step 1: Try Bray Curtis 

First, convert to community matrix.

```{r data to matrix}
prop.reads <- full.proportion.reads[,4:6] # only eDNA proportion values
prop.mat<- t(prop.reads) # columns = Hashes, rows = PCR #, values = nReads proportions
```



```{r BCD between PCR runs across samples}
vegdist(prop.mat, method="bray", na.rm = T)
```

All yielding pretty similar BC values (~0.41). But I'm worried that we might be glossing over some variability between bottles (ie, variation b/w PCR rep 1, 2 and 3 might be different based on the sample). Let's try getting PCR BC dissimilarity by sample. 

```{r BCD between PCR runs between samples}

 #bioVector = list of unique biological replicate IDs, indexCols = col numbers of eDNA proportion values (by PCR #). df needs to have 3 columns corresponding to eDNA proportions of different PCR runs

bray_by_sample <- function(df, bioVector, indexCols){  
  
    bray.mat <- matrix(NA, nrow = length(bioVector), ncol = 3, dimnames = list(bioVector, c(1.2,2.3,1.3))) # different PCR rep combinations
  
  for(i in 1:length(bioVector)){
    sample.tib <- df[df$bio == bioVector[i],indexCols]
    bray.tib <- t(sample.tib)
    bray.results <- as.matrix(vegdist(bray.tib, method="bray", na.rm = T))
    bray.mat[i,] <- c(bray.results[1,2], bray.results[3,2], bray.results[1,3]) # PCR 1&2, 2&3, 1&3
  }
    return(bray.mat)
}

bray_by_sample(full.proportion.reads, unique(full.proportion.reads$bio), c(4:6))

```

Also, I don't want to overfocus on variability between samples, because that's not what we're interested in. So maybe the chunk above is totally obsolete. However, I wonder if doing a permanova between the three *distributions* of BCDs (separated by sample) is more informative than just looking at 3 BCD values (for each PCR combination, across samples). ????

First, let's visualize how BCD distributions look in different PCR runs. 

```{r Visualize BCD distributions}

bray.output <-   bray_by_sample(full.proportion.reads, unique(full.proportion.reads$bio), c(4:6))

par(mfrow = c(1, 3))
hist(bray.output[,1], main="PCRs 1 & 2", xlab="BCD", ylim=c(0,60), col = "orange") 
hist(bray.output[,2], main="PCRs 2 & 3", xlab="BCD", ylim=c(0,60), col = "orange") 
hist(bray.output[,3], main="PCRs 1 & 3", xlab="BCD", ylim=c(0,60), col = "orange")
```

Now, let's compare distributions. We have to change the matrix around a bit.

```{r PERMANOVA on 3 BCD distributions}

bray.perm <- melt(as.data.frame(bray.output))
colnames(bray.perm) <- c("PCR.Combo", "BCD")
adonis(BCD ~ PCR.Combo,data = bray.perm, permutations = 9999)
```

Hm p value of 0.397. So at very first glance, reads don't differ that much between PCR replicates. (p < 0.05 = sig dissimilarity) 

*Notes on permanova + adonis:* 
- "Permutational multivariate analysis of variance (PERMANOVA) is a non-parametric multivariate statistical test. It is used to compare groups of objects and test the null hypothesis that the centroids and dispersion of the groups as defined by measure space are equivalent for all groups." 
- So we want to see how the DISSIMILARITY between samples in PCR Rep 1&2 compares to the DISSIMILARITY between samples in PCR Rep 1&3, etc. 
- adonis tests for homogeneity of dispersion among groups


## Step 2: Try Beta Diversity (looking at the number of unique hashes/taxa in samples between PCR reps)

So beta diversity looks at total # of unique species in each sample/PCR replicate. Let's repeat what we did for the BCD (dissimilarity by sample between PCR reps)

```{r Beta Div test & visualize }
# again, betadiver wants PCR reps (equate to site) as rows, and Hashes (equate to species) as columns
# in betadiver(), we're using method 1 (formula = (b+c)/2)

beta_by_sample <- function(df, bioVector, indexCols){  
  
    beta.mat <- matrix(NA, nrow = length(bioVector), ncol = 3, dimnames = list(bioVector, c(1.2,2.3,1.3))) # different PCR rep combinations
  
  for(i in 1:length(bioVector)){
    sample.tib <- df[df$bio == bioVector[i],indexCols]
    beta.tib <- t(sample.tib)
    beta.results <- as.matrix(betadiver(beta.tib, method = 1))
    beta.mat[i,] <- c(beta.results[1,2], beta.results[3,2], beta.results[1,3]) # PCR 1&2, 2&3, 1&3
  }
    return(beta.mat)
}

beta.output<- beta_by_sample(full.proportion.reads, unique(full.proportion.reads$bio), c(4:6))

par(mfrow = c(1, 3))
hist(beta.output[,1], main="PCRs 1 & 2", xlab="BDiv", col = "red")
hist(beta.output[,2], main="PCRs 2 & 3", xlab="BDiv", col = "red")
hist(beta.output[,3], main="PCRs 1 & 3", xlab="BDiv", col = "red")
```

(!!) I want to talk about this more...I'm not confident in how to interpret the beta diversity output. I think 1 represents high dissimilarity though? In Hash richness within a sample between PCR reps. 


```{r permanova on bdivs}
beta.perm <- melt(as.data.frame(beta.output))
colnames(beta.perm) <- c("PCR.Combo", "BetaDiv")
adonis(BetaDiv ~ PCR.Combo,data = beta.perm, permutations = 9999) # p=0.88
```


## Step 3: Try Jaccard Similarity (bias towards rare sp)

*Notes on jaccard* 
- "measures the similarity between two sets of data. It can range from 0 to 1. The higher the number, the more similar the two sets of data."
- vegdist jaccard (same as earlier BCD) wants taxa/hashes as columns.

Pretty much just repeating what I did for the BCD above. 

```{r Jaccard between PCR runs between samples and visualize}

jac_by_sample <- function(df, bioVector, indexCols){  
  
  jac.mat <- matrix(NA, nrow = length(bioVector), ncol = 3, dimnames = list(bioVector, c(1.2,2.3,1.3))) # different PCR rep combinations
  
  for(i in 1:length(bioVector)){
    sample.tib <- df[df$bio == bioVector[i],indexCols]
    jac.tib <- t(sample.tib)
    jac.results <- as.matrix(vegdist(jac.tib, method="jaccard"))
    jac.mat[i,] <- c(jac.results[1,2], jac.results[3,2], jac.results[1,3]) # PCR 1&2, 2&3, 1&3
  }
    return(jac.mat)
}

jac.output <- jac_by_sample(full.proportion.reads, unique(full.proportion.reads$bio), c(4:6))

par(mfrow = c(1, 3))
hist(jac.output[,1], main="PCRs 1 & 2", xlab="Jaccard", col = "purple")
hist(jac.output[,2], main="PCRs 2 & 3", xlab="Jaccard", col = "purple")
hist(jac.output[,3], main="PCRs 1 & 3", xlab="Jaccard", col = "purple")
```

```{r permanova on jaccard}
jac.perm <- melt(as.data.frame(jac.output))
colnames(jac.perm) <- c("PCR.Combo", "JaccardDiv")
adonis(JaccardDiv ~ PCR.Combo,data = jac.perm, permutations = 9999) # p=0.411
```

So thus far, permanovas on Bray Curtis, Beta Diversity, and Jaccard distance have shown that PCR replicates don't produce significantly different communities within biological samples. 

Another note: across strategies, 14% of all dissimilarity values were 1. 