---
title: "ASV_var_negbin"
author: "Ryan Kelly, Helen Casendino, Joe Duprey, Kai Vennemann"
date: "10/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ASV variation using Negative Binomial Parameters

Uses Ryan's code to fit PCR reads for each ASV (usually 3) to a negative binomial distribution. Estimates values for mu and phi in the process.

```{r dependencies}
library(tidyverse)
library(stringi)
library(fitdistrplus)
library(ggpubr)
library(here)
```

Read in OA data.

```{r read in data}
all_data <- read_csv(here("data/all_data.csv"))
```

Function to get the variance of a negbin distribution, given the parameters mu and phi.

```{r create varNB function}
varNB <- function(mu, phi) {
  mu + ((mu^2) / phi)
} # calculates variance
```

Function that uses fitdist() to find the most likely parameters mu and phi (which R calls "size" in this case), given some data vector.

```{r create getNB function}
getNB <- function(x) {
  require(fitdistrplus)

  return(
    (fitdist(unlist(x), "nbinom"))$estimate
  )
} # estimates mu and phi
```

Then code below takes a data set, groups by species and biological sample, and then fits a negative binomial distribution to each vector of 3 observations for each species at each site. Then it plots the result.

```{r create estimates_pars_byHash function, warning=FALSE}
# Below: estimates parameters for each Hash (2-3 PCR reads/Hash)
# df needs cols: Miseq_run, bio (bottle, ie PO2017A), Hash, and nReads
# rep_type is either "technical" for PCR variation or "biological" for bottle var
# spits out list of:
# 1) df of mu, estimated variance and phi, and original reads for each ASV
#   and 2) plot of mu vs. estimated variance and mu vs. phi


estimates_pars_byHash <- function(df, rep_type) {
  if (rep_type == "technical") {
    grouping_df <- df %>%
      mutate(bottle = paste(site, bio, sep = "_"))
      filter(bottle %in% unique(df$bottle)) %>%
      group_by(bottle, hash)
  }

  if (rep_type == "biological") {
    grouping_df <- df %>%
      filter(site %in% unique(df$site)) %>% # WHAT IS THIS LINE DOING?
      group_by(site, hash)
  }

  list_NBpars <- grouping_df %>%
    filter(sum(reads) > 0) %>%
    filter(n() > 1) %>%
    dplyr::select(reads) %>%
    nest() %>%
    mutate(mlmodels = map(data, getNB))

  df_NBpars <- list_NBpars$mlmodels %>%
    as.data.frame() %>%
    t() %>%
    as_tibble() %>%
    rename(phi = size) %>%
    mutate(estVariance = varNB(mu, phi)) %>%
    mutate(Reads = list_NBpars$data) %>%
    pivot_longer(-c(mu, Reads))

  NBpars_plot <- df_NBpars %>%
    ggplot(aes(x = log(mu), y = log(value))) +
    geom_point() +
    facet_wrap(~name)

  return(list(df_NBpars, NBpars_plot))
}

NB_output <- estimates_pars_byHash(all_data, rep_type = "biological")
```

Helen: If the curve of mu vs. variance is dependent on phi, we also want to keep track of the slope of mu vs. variance, to compare to the slope of other kinds of replicates?

