---
title: "differences.ASV.var"
author: "Helen Casendino, Joe Duprey"
date: "9/28/2021"
output: html_document
---

**Goal:** To assign distributions of variation, in units of arithmetic differences (think of better word?), to technical (PCRs) and biological (bottles) sources at the ASV level.

**Strategy:** 
1) To assign technical PCR variation, get three pairwise differences between PCR replicates for each ASV (basically, p_PCR1 - p_PCR2, p_PCR1 - p_PCR3, and p_PCR2 - p_PCR3). For now, let's take the absolute value of each difference.
2) To assign random biological variation, take pairwise differences in the same fashion across all bioPCRs for each ASV (A1,B2,C1 etc). However, don't take pairwise differences *within* bottles (A1,A2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r dependencies, include=FALSE}
library(tidyverse)
library(stringi)
```

## Part 1: Read in data 

```{r read in data}
PCR_props <- read.csv("../../data/all_data_clean.csv")
```

## Part 2: PCR variation (within bottles)

```{r pairwise differences between PCRs}
calc_ASV_PCR_diffs <- function(df) {
  df <- df %>% 
    ungroup() %>%
    select(bio, tech, hash, prop)
  
  bottles <- unique(df$bio)
  hashes <- unique(df$hash)
  diffs_upper_bound <- choose(max(df$tech), 2) * length(bottles) * length(hashes)
  diffs <- rep(NA, diffs_upper_bound)  # Vector to store diffs
  i <- 1
  
  # Iterate over each bottle and calculate diffs among PCRs
  for (bottle in bottles) {
    bottle_data <- df %>% 
      filter(bio == bottle) %>% 
      pivot_wider(names_from = tech, values_from = prop, values_fill = 0) %>% 
      select(matches("\\d{1, }"))  # Columns that have numbers as names (i.e., represent PCRs)
    
    pairs <- choose(ncol(bottle_data), 2)
    
    # Iterating over each hash for this particular bottle
    for(j in 1:nrow(bottle_data)) {
      diffs[i:(i + pairs - 1)] <- c(dist(as.numeric(bottle_data[j, ])))  # Pairwise diffs for each hash
      i <- i + pairs
    }
  }
  
  diffs <- as.numeric(na.omit(diffs))
  
  # Sanity check
  print("ASV PCR Differences:")
  print(paste("Mean diff:", mean(diffs), sep = " "))
  print(paste("Proportion of diffs that equal zero:", mean(diffs == 0), sep = " "))
  
  # Plot
  hist(
    diffs,
    col = viridis::plasma(3, 0.4, 0.7),
    main = "ASV PCR Variation",
    xlab = "Pairwise Differences"
  )
  
  hist(
    log(diffs),
    col = viridis::plasma(3, 0.4, 0.7),
    main = "ASV PCR Variation (Logarithmic)",
    xlab = "Logarithm of Pairwise Differences"
  )

  diffs
}

ASV_PCR_diffs <- calc_ASV_PCR_diffs(PCR_props)  # mean = 0.0018
```

Thought train: 
Hm...so since the histogram is just going to be skewed no matter what, and we will still have 0s, I wonder if there is much benefit to normalizing? Also the log data looks good but it can't handle the 0s so000...we need to find the right statistical distribution...Since they're positive values b/w 0 and 1, beta? Does beta accept 0s? hm. 

Also, there's an absurdly large proportion of zeros; maybe we should deal with these???

## Part 3: Bottle variation (within samples)

As always, we need to include this chunk to nest the data from each sample (each sample has a different # of bottles)..

```{r data wrangling, not taking PCR mean}
long.props <- trip.proportion.reads %>%
  mutate(sample = gsub(".{1}$", "", bio)) %>%
  mutate(bottle = stri_sub(bio, -1)) %>%
  pivot_longer(cols = c(PCR1_prop, PCR2_prop, PCR3_prop), names_to = "PCR", values_to = "Proportion")

get.PCRn <- long.props %>%
  mutate("PCRn" = stri_sub(PCR, 4, -6)) %>%
  select(-c(bio, PCR))

nested.props <- get.PCRn %>%
  unite(bottle, PCRn, col = "bio.PCR", sep = ".") %>%
  nest(data = c(bio.PCR, Hash, Proportion))

for (i in 1:length(nested.props$sample)) {
  nested.props$data[[i]] <- nested.props$data[[i]] %>%
    group_by(bio.PCR, Hash) %>%
    summarise(Proportion = mean(Proportion)) %>%
    pivot_wider(names_from = bio.PCR, values_from = Proportion, values_fill = 0)

  print(i)
}
```

```{r bottle ASV differences function}
diffs.BOTTLE <- function(list, SampleVector) {
  diffs <- "start"


  # Getting rid of samples with only one bottle
  onebottle <- rep(NA, length = 100) # we will put the "list" indices of replicates with only one bottle here

  for (i in 1:length(SampleVector)) {
    a <- list[[i]]
    if (((ncol(a) - 1) / 3) == 1) { # replicates with only 1 bottle
      onebottle[i] <- i # save index
      SampleVector <- SampleVector[-i] # remove index from sample vector
    }
  }
  onebottle <- onebottle[!is.na(onebottle)] # remove NAs
  list <- list[-onebottle] # remove the one bottle elements from list


  for (j in 1:length(SampleVector)) {
    print(j)
    df <- list[[j]]
    sub <- df %>% select(!c(Hash))

    for (i in 1:length(sub[, 1])) {
      x <- as.numeric(sub[i, ]) # single row = single ASV

      for (p in 1:15) { # this only collects the differences between bottles, not within bottles (explanation below)
        if (p == 1 | p == 4 | p == 7 | p == 10 | p == 13) {
          if (p < (length(x) - 2)) {
            diffs <- c(diffs, (abs(x[p] - x[(p + 3):length(x)])))
          }
        }
        if (p == 1 | p == 4 | p == 7 | p == 10 | p == 13) {
          if (p < (length(x) - 2)) {
            diffs <- c(diffs, (abs(x[p] - x[(p + 3):length(x)])))
          }
        }
        if (p == 1 | p == 4 | p == 7 | p == 10 | p == 13) {
          if (p < (length(x) - 2)) {
            diffs <- c(diffs, (abs(x[p] - x[(p + 3):length(x)])))
          }
        }
      }
    }
  }
  hist <- hist(as.numeric(diffs[-1]), col = viridis::viridis(3, 0.4, 0.7), main = "Bottle Variation", xlab = "Pairwise Differences")

  return(as.numeric(diffs[-1]))
  return(hist)
}

bottle.output <- diffs.BOTTLE(nested.props$data, nested.props$sample)
# mean = 0.0013
``` 

About the p loop: So we need to get differences between bottles, not within the same bottle (so A1-B1, not A1-A2). And many samples have different numbers of bottles, so we need a way to tell the code which indices, within the row vector of proportions, to compare. There will be at most 18 indices (6 bottles). The (abs(x[p] - x[(p+3):length(x)]) take the index value corresponding to a replicate (ie, A1, or x[1]), and compares it to all other replicates not in the same bottle (x[1] - x[4:length(x)]). We need the if(p < (length(x) - 2)) so that the loop doesn't take differences within the last bottle in each sample...for nitpicky reasons not worth going into complete depth here. 
