---
title: "pres.abs.agreement"
author: "Helen Casendino, Joe Duprey"
date: "10/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
```

How does presence/absence agreement - consistency of presence/absence detections across PCR replicates
change as read proportion increases/decreases. Based on Kai's question.

```{r}
# load clean PCR rep data
trip.proportion.reads <- read.csv('../../data/PCR.proportions.clean.csv')

# get mean proportion across PCR replicates, log(mean_prop) to make cutoffs easier? 
PA.agree <- trip.proportion.reads %>%
  mutate(mean_prop = (PCR1_prop + PCR2_prop + PCR3_prop) / 3) %>%
  mutate(log_mean_prop = log(mean_prop))

# column to mark when there are 3/3 detections/presence 
PA.agree <- PA.agree %>%
  mutate(
    agree = case_when(
      PA.agree$PCR1_prop > 0 & PA.agree$PCR2_prop > 0 & PA.agree$PCR3_prop > 0 ~ 1,
      TRUE ~ 0) #confusing syntax but this marks all other cols as 0 
    )
 
```

Now we loop through log_mean_prop and see how the proportion that agree change

```{r}
# get the min and max values 
print(min(PA.agree$log_mean_prop))
print(max(PA.agree$log_mean_prop))

# sequence - we should probably try this without log() too to make it more interpretable? 
prop.seq = seq(from = -13.35, to = -0.5, by = 0.1)

# summary table of all rows 
sumtable <- table(PA.agree$agree)
print(sumtable)


# lists to store values for plotting 
prop.ag.list <- vector(length=length(prop.seq))
j <- 1 

# loop through cutoff points
for (i in prop.seq){
  print(i)
  
  # filter the df 
  filtered.df <- PA.agree %>%
    filter(PA.agree$log_mean_prop < i)
  
  # calculate the proportion where 3/3 agree as the log(mean_cutoff) increases 
  prop.agreement <- sum(filtered.df$agree) / nrow(filtered.df)
  prop.ag.list[j] <- prop.agreement # store the resulting proportion
  
  # print the filtered df length for sanity check 
  print(nrow(filtered.df))
  
  # joe's index variable - sleeker way to do this?? 
  j <- j + 1 
  
}
```

```{r}
#plot the results
plot.df = data_frame(prop.seq, prop.ag.list)
plot.df$mean_log_prop <- prop.seq

plot(plot.df$prop.seq, plot.df$prop.ag.list, xlab = 'cutoff point - log(mean proportion)', ylab = 'proportion agreement of Presence/Absence')
```
The moving cutoff point is hard for me to interpret because the sample size is always going up -
lets try binning log(mean_read_prop) instead

```{r}
# bin sequence, 24 bins to start
bin.seq = seq(from = -13.35, to = -0.5, length.out = 25)
print(bin.seq)

nbins <- length(bin.seq) - 1
agree.values <- vector(length=nbins)

for (i in 1:nbins){
  print(bin.seq[i])
  print(bin.seq[i+1])
  print('-----')
  
  one.bin <- PA.agree %>%
    filter(PA.agree$log_mean_prop >= bin.seq[i] & PA.agree$log_mean_prop < bin.seq[i+1])
  
  bin.agreement <- sum(one.bin$agree) / nrow(one.bin)
  agree.values[i] <- bin.agreement
  
  
  
}

plot(agree.values)
```


